---
title: "PHI Applied Research Fellows 2021\\ Intro to Demography"
author: "Jessica Godwin"
date: "June 23, 2021"
output: 
  beamer_presentation:
    theme: "Dresden"
header-includes: 
  - \addtobeamertemplate{title page}{\includegraphics[width=1.5cm]{W-Logo_Purple_RGB} \hfill \includegraphics[width=1.5cm]{csdelogo}}{}
classoption: "aspectratio=169"
urlcolor: blue
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "~/Dropbox/PHI2021/Slides/")
if(FALSE){
  setwd("~/Dropbox/PHI2021/Slides/")
}

sandArt <- function(popsize, popcontains,
                    x.pts, x.lims,
                    colors, color.cuts,
                    plot.dat,
                    colors.box = 0000){
  plot(NA,
       xlab = "",
       ylab = paste0("Proportion of ",
                     popcontains),
       xlim = c(0, popsize),
       ylim = c(0, popsize),
       xaxt = 'n', yaxt = 'n',
       axes = FALSE)
  
  if(sum(x.lims != range(x.pts)) == 0){
    fake.x.pts <- popsize*((x.pts-x.pts[1])/
                             (x.pts[length(x.pts)] - x.pts[1]))
  }else{
    fake.x.pts <- popsize*((x.pts-x.lims[1])/
                             (x.lims[2] - x.lims[1]))
  }
  
  
  for(lower in color.cuts[-length(color.cuts)]){
    idx <- match(lower, color.cuts)
    n.by.time <- apply(as.matrix(plot.dat), 2, function(value){
      sum(value >= lower & 
            value < color.cuts[idx + 1],
          na.rm = TRUE)
    })
    
    if(idx == 1){
      y1 <- rep(0,length(fake.x.pts))
      y2 <- rep(0,length(fake.x.pts))
    }else{
        y1 <- y2 
    }
    
    if(idx != length(color.cuts) - 1){
      y2 <- y1 + n.by.time
    }else{
      y2 <- rep(popsize, length(x.pts))
    }                 
    
    # print(paste0("\n", y1, "--", y2))
    # print(paste0("\n Age: ", lower, "\n"))
    # print(paste(n.by.time, collapse = "  "))
    polygon(x = c(fake.x.pts,
                  rev(fake.x.pts)),
            y = c(y1, rev(y2)),
            col = colors[idx],
            border = FALSE)
    
    if(lower %in% colors.box){
      lines(x = c(fake.x.pts,
                  rev(fake.x.pts)),
            y = c(y1, rev(y2)),
            col = colors[idx], lwd = 3)
    }
  }
  
  axis(2, at = popsize*seq(0,1,.2),
       labels = seq(0, 1, .2))
}
```


```{r libs, include = FALSE}
library(wesanderson)
library(xtable)
library(dplyr)
library(tidyr)
library(scales)
library(wpp2019)
library(viridisLite)
```

# What is Demography?


\pause 
$$P_{t+1} = P_t + B_t - D_t + IM_t - OM_t$$

* Fertility, mortality, migration, population size
- The balancing equation
* How these processes work together in a population
* Break it all down by age and sex and ...

# Why is Demography?

\pause
* Understand a population's make up today
* Targeted intervention
* Projections allow planning for future population
* Historical demography
* Social demography

# The main textbooks

*  Demography: Measuring and Modeling Population Processes
- Samuel Preston, Patrick Heuveline, Michell Guillot
*  Essential Demographic Methods 
- Kenneth Wachter
*  [Tools for Demographic Estimation](https://demographicestimation.iussp.org/content/introduction-tools-demographic-estimation)
- IUSSP (many prominent world demographers)

# The Demographic Transistion Theory

 * Mortality begins to decrease $\rightarrow$
 * Fertility begins to decrease $\rightarrow$
 * In countries that have already experienced the transition, to below "replacement level" $\approx$ 2.1 children per parents $\rightarrow$ rise to and fluctuation around this point
  * behind model in WPP (Alkema et al., 2011)
  * Is this true in places where TFR remains high? Is all "high" fertility a result of unmet family planning needs?
 * [This really succinct amazing graphic on Wikipedia](https://en.wikipedia.org/wiki/Demographic_transition#/media/File:Demographic-TransitionOWID.png)

# N Africa & West Asia: 1950-1955 to 2015-2020
```{r, dtNAWA,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
data(UNlocations)
data(e0F)
data(tfr)
data(pop)

# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 1
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx] == codes[code.idx])]
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```


# Sub-Saharan Africa: 1950-1955 to 2015-2020
```{r, dtSSA,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
data("UNlocations")
data(e0F)
data(tfr)
data(pop)

# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 2
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx] == codes[code.idx])]
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  n.countries <- nrow(plot.tmp)
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```


# Central & South Asia: 1950-1955 to 2015-2020
```{r, dtCSA,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 947 SSA
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 3
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```  

# East & South-Eastern Asia: 1950-1955 to 2015-2020
```{r, dtESEA,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 947 SSA
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 4
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx] == codes[code.idx])]
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```  

# Latin America \& the Caribbean: 1950-1955 to 2015-2020
```{r, dtLAC,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 947 SSA
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 5
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx] == codes[code.idx])]
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)

  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```  

# Australia, New Zealand, Oceania: 1950-1955 to 2015-2020
```{r, dtOce,eval=TRUE,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 947 SSA
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 6:7
# Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx[1]], names(UNlocations))
  col.idx <- c(col.idx, grep(codes[code.idx[2]], names(UNlocations)))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx[1]] == codes[code.idx[1]])]
  sub.codes <- c(sub.codes,
                 UNlocations$country_code[which(UNlocations[,col.idx[2]] == codes[code.idx[2]])])
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```  

# Europe \& North America: 1950-1955 to 2015-2020
```{r, dtENA,fig.height=3.5,message = FALSE, warning = FALSE, results='asis'}
# 1 "World"
# 903 Africa
# 935 Asia
# 908 Europe
# 904 LA&Car
# 905 North America
# 909 Oceania
# 1833 NA & W Asia
# 947 SSA
# 921 C & S Asia
# 1832 E & SE Asia
# 1830 LA & Car
#  927 Aus/NZ
# 1835 Oceania
# 1829 E & NA
countries <- UNlocations$country_code[which(UNlocations$location_type == 4)]
codes <- c(1833,947, 921, 1832, 1830, 927, 1835, 1829)
x.lim <- c(1950,2020)
plot.years <- seq(1952, 2017, 5)
e0.cuts <- c(15, 25, 45, 55, 65,
             70, 75, 80, 85, 90)
cols <- wes_palette("Zissou1", 5,type = c("discrete"))[c(1,3,5)]
code.idx <- 8
Area <- UNlocations$name[UNlocations$country_code == codes[code.idx]]
{
  col.idx <- grep(codes[code.idx], names(UNlocations))
  sub.codes <- UNlocations$country_code[which(UNlocations[,col.idx] == codes[code.idx])]
  
  par(lend = 1,
      mar = c(2,2,2,2) + 0.1)
  layout(matrix(c(1,2,3), nrow = 1, byrow = TRUE),
         widths = c(1,1,.33), heights = 1)
  plot.tmp <- e0F %>% filter(country_code %in% sub.codes)
  n.countries <- nrow(plot.tmp)
  e0.cols <- alpha(cols[3],
                   seq(1,.15, length.out = length(e0.cuts)-1))
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = e0.cols, 
          color.cuts = e0.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]))
  axis(1, at = n.countries*((seq(1952, 2022, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10))
  title(main = expression("Life Expectancy, e"[0]))
  plot.tmp <- tfr %>% filter(country_code %in% sub.codes)
  tfr.cols <- rev(alpha(cols[1],
                        seq(1,.15, length.out = length(e0.cuts)-1)))
  tfr.cuts <- c(0.75, 1, 1.5, 2.1, 2.5,
                3, 4, 5, 6, 9)
  sandArt(popsize = n.countries,
          popcontains = "Countries",
          x.pts =  plot.years,
          x.lims = x.lim,
          colors = tfr.cols, 
          color.cuts = tfr.cuts,
          plot.dat = as.matrix(plot.tmp[,-c(1:2,ncol(plot.tmp))]),
          colors.box = 2.1)
  axis(1, at = n.countries*((seq(1950, 2020, 10)-x.lim[1])/(x.lim[2]-x.lim[1])), 
       labels = seq(1950,2020,10)) 
  title(main = "Total fertility rate (TFR)")
  plot(NA,
       xlim = c(0,1),
       ylim = c(0,1),
       xlab = "",
       ylab = "",
       axes = FALSE)
  legend.all <- c(e0.cuts, tfr.cuts)
  legend.order <- legend.all
  order <- 0
  for(i in 1:9){
    order <- order + 1
    legend.order[i] <- order
    order <- order + 1
    legend.order[i + 9] <- order
  }
  legend("center",
         fill = c(e0.cols,
                  tfr.cols),
         ncol = 2,
         cex = 1,
         box.col = 'white',
         legend = c(e0.cuts[-1],
                    tfr.cuts[-1]),
         border = FALSE)
}  
```  

# Population Growth by Continent

```{r continent, fig.align='center',fig.height = 3, fig.width = 9}
y.lim <- c(0, 2500000)
cols <- rainbow(length(codes))


par(mfrow = c(1,3),
    mar = c(2,2,2,2))
plot(NA,
     xlim = x.lim,
     ylim = y.lim,
     xaxt = 'n',
     yaxt = 'n',
     axes = FALSE,
     xlab = "Year",
     ylab = "Population in 1000s")
axis(1, seq(1950,2020,10))
axis(2, seq(0, 2500000, length.out = 6))
title(main = "Population")
plot.tmp <- pop %>% filter(country_code %in% codes)


for(code in codes){
  tmp <- plot.tmp[plot.tmp$country_code == code,]
  lines(plot.years,
        tmp[-c(1:2,ncol(tmp))],
        lwd = 2, col = cols[match(code,codes)],
        type = 'b')
}


legend('topleft',
       lty = 1, lwd = 2,
       col = cols, pch = 1,
       bty = 'n', cex = .8,
       legend = c("N Africa & W Asia",
                  "Sub-Saharan Africa",
                  "Central & S Asia",
                  "Latin America & Carribean",
                  "E & SE Asia",
                  "Australia/NZ",
                  "Oceania",
                  "Europe & N America"))

plot.tmp <- e0F %>% filter(country_code %in% codes)
y.lim <- c(35,85)
plot(NA,
     xlim = x.lim,
     ylim = y.lim,
     xaxt = 'n',
     yaxt = 'n',
     axes = FALSE,
     xlab = "Year",
     ylab = expression(paste0("Life expectancy, ",e[0])))
axis(1, seq(1950,2020,10))
axis(2, seq(35,85, 10))
title(main = "Mortality")

for(code in codes){
  tmp <- plot.tmp[plot.tmp$country_code == code,]
  lines(plot.years,
        tmp[-c(1:2,ncol(tmp))],
        lwd = 2, col = cols[match(code,codes)],
        type = 'b')
}



plot.tmp <- tfr %>% filter(country_code %in% codes)
y.lim <- c(0.75,7)
plot(NA,
     xlim = x.lim,
     ylim = y.lim,
     xaxt = 'n',
     yaxt = 'n',
     axes = FALSE,
     xlab = "Year",
     ylab = "Total fertility rate (TFR)")
axis(1, seq(1950,2020,10))
axis(2, c(1, 2.1, 3, 5, 7))
title(main = "Fertility")
for(code in codes){
  tmp <- plot.tmp[plot.tmp$country_code == code,]
  lines(plot.years,
        tmp[-c(1:2,ncol(tmp))],
        lwd = 2, col = cols[match(code,codes)],
        type = 'b')
}


```

# Population Pyramids: Europe, 2015

```{r pyr1, fig.align='center',fig.height=4,fig.width=4, message=FALSE}
library(bayesPop)
data(popF)
data("popM")
popF.long <- popF %>% pivot_longer(contains(c("5", "0")),
                                    names_to = "Year",
                                    values_to = "female")

ages <- unique(popM$age)
pyr.data <- popM %>% pivot_longer(contains(c("5", "0")),
                                    names_to = "Year",
                                    values_to = "male") %>%
  select(-name) %>%
  filter(Year == 2015) %>%
  left_join(popF.long, by = c("country_code",
                               "Year", "age"))

pyr.tmp <-pyr.data %>%
  filter(name == "EUROPE") %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages

pyr <- get.bPop.pyramid(pyr.tmp, legend = "Europe")
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]))

```
# Population Pyramids: Latin America \& Caribbean, 2015

```{r pyr2, fig.align='center',fig.height=4,fig.width=4}


pyr.tmp <-pyr.data %>%
  filter(name == "LATIN AMERICA AND THE CARIBBEAN") %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages

pyr <- get.bPop.pyramid(pyr.tmp, legend = "Latin America/Caribbean")
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]))
```

# Population Pyramids: Asia, 2015
```{r pyr3, fig.align='center',fig.height=3,fig.width=6}
pyr.tmp <-pyr.data %>%
  filter(country_code == 921) %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages
par(mfrow=c(1,2))
pyr <- get.bPop.pyramid(pyr.tmp)
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]), main = "South-Central Asia")

pyr.tmp <-pyr.data %>%
  filter(name == "Eastern Asia") %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages

pyr <- get.bPop.pyramid(pyr.tmp)
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]), main = "Eastern Asia")
```

# Population Pyramids: Africa, 2015

```{r pyr4, fig.align='center',fig.height=3,fig.width=6}

pyr.tmp <- pyr.data %>%
  filter(name == "Northern Africa") %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages
par(mfrow=c(1,2))
pyr <- get.bPop.pyramid(pyr.tmp)
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]),
     main = "North Africa")

pyr.tmp <- pyr.data %>%
  filter(country_code == 947) %>%
  select(c("male", "female")) %>%
  as.matrix()
rownames(pyr.tmp) <- ages

pyr <- get.bPop.pyramid(pyr.tmp)
plot(pyr, pyr1.par = list(col = tfr.cols[9], 
                          border = tfr.cols[9]),
     main = "Sub-Saharan Africa")

```


# Population Pyramids: WA, 2011
 * [Over time from WA OFM](https://ofm.wa.gov/washington-data-research/statewide-data/washington-trends/population-changes/distribution-washington-population-age-and-gender)
```{r pyrWA,message=FALSE,fig.height=4.5}
data <- read.table(file.path(find.package("bayesPop"), "ex-data", "popestimates_WAKing.txt"), 
    header=TRUE, row.names=1)
# extract data for two pyramids and put it into the right format
WA <- data[,c("WA.male", "WA.female")]; colnames(WA) <- c("male", "female")
King <- data[,c("King.male", "King.female")]; colnames(King) <- c("male", "female")
pyramid <- get.bPop.pyramid(list(WA, King), legend=c("Washington", "King"))
plot(pyramid,
     pyr1.par = list(col = alpha(tfr.cols[9],0.7),
                  border = tfr.cols[9]),
     pyr2.par = list(col = e0.cols[1],
                  border = e0.cols[1]))


```


# Mortality: Age-specific Mortality

```{r mxfig, echo = FALSE, fig.align='center', fig.height=4,fig.width=4}
data(mxM)
data(mxF)

countries <- c("United States of America", "Zimbabwe")
ages <- unique(mxM$age)

sex.cols <- wes_palette("Zissou1", n = 5)[c(1,3)]

years <- c("1995-2000", "2005-2010", "2015-2020")
ylim <- range(log(mxM[mxM$name %in% countries, years])) + .1*c(-1,1)
y.axis.num <- seq(-10,0,2)
y.axis.label <- round(exp(y.axis.num)*1000, 0)

par(mfrow = c(1,1), lend = 1)
for(year in years[1]){
  plot(NA,
       xlim = range(ages),
       ylim = ylim,
       yaxt = 'n',
       xlab = "Age",
       ylab = "Mortality rate (per 1000)",
       main =  year)
  axis(2, at = y.axis.num, labels = y.axis.label)
  
  lines(ages, log(mxM[mxM$name == countries[1], year]),
        type = 'b',
        col = alpha(sex.cols[2], 0.5),
        pch = 19)
  lines(ages, log(mxF[mxF$name == countries[1], year]),
        type = 'b',
        col = alpha(sex.cols[1], 0.5),
        pch = 19)
  lines(ages, log(mxM[mxM$name == countries[2], year]),
        type = 'b',
        col = alpha(sex.cols[2], 1),
        pch = 19)
  lines(ages, log(mxF[mxF$name == countries[2], year]),
        type = 'b',
        col = alpha(sex.cols[1], 1),
        pch = 19)
  legend('bottomright', bty = 'n',
         lty = 1, pch = 19, cex = 0.75,
         col = c(alpha(sex.cols[2:1],0.5),
                 sex.cols[2:1]),
         legend = c("US: Male", "US: Female",
                    "Zimbabwe: Male", "Zimbabwe: Female"))
}

```

# Mortality: Age-specific Mortality

```{r mxfig2, echo = FALSE, fig.align='center', fig.height=3.5, fig.width=3.5}
data(mxM)
data(mxF)

countries <- c("United States of America", "Zimbabwe")
ages <- unique(mxM$age)

sex.cols <- wes_palette("Zissou1", n = 5)[c(1,3)]

years <- c("1995-2000", "2005-2010", "2015-2020")
ylim <- range(log(mxM[mxM$name %in% countries, years])) + .1*c(-1,1)
y.axis.num <- seq(-10,0,2)
y.axis.label <- round(exp(y.axis.num)*1000, 0)

par(mfrow = c(1,1), lend = 1)
for(year in years[3]){
  plot(NA,
       xlim = range(ages),
       ylim = ylim,
       yaxt = 'n',
       xlab = "Age",
       ylab = "Mortality rate (per 1000)",
       main =  year)
  axis(2, at = y.axis.num, labels = y.axis.label)
  
  lines(ages, log(mxM[mxM$name == countries[1], year]),
        type = 'b',
        col = alpha(sex.cols[2], 0.5),
        pch = 19)
  lines(ages, log(mxF[mxF$name == countries[1], year]),
        type = 'b',
        col = alpha(sex.cols[1], 0.5),
        pch = 19)
  lines(ages, log(mxM[mxM$name == countries[2], year]),
        type = 'b',
        col = alpha(sex.cols[2], 1),
        pch = 19)
  lines(ages, log(mxF[mxF$name == countries[2], year]),
        type = 'b',
        col = alpha(sex.cols[1], 1),
        pch = 19)
  legend('bottomright', bty = 'n',
         lty = 1, pch = 19, cex = 0.75,
         col = c(alpha(sex.cols[2:1],0.5),
                 sex.cols[2:1]),
         legend = c("US: Male", "US: Female",
                    "Zimbabwe: Male", "Zimbabwe: Female"))
}

```

# The Lexis Diagram: Age-Period-Cohort

\begin{figure}[h]
\centering
\includegraphics[height=0.8\textheight]{LexisLifeline.pdf}
\end{figure}


# The Lexis Diagram: Age-Period-Cohort

\begin{figure}[h]
\centering
\includegraphics[height=0.8\textheight]{LexisPopSummary.pdf}
\end{figure}

# The Life Table: Age-Period-Cohort
\begin{figure}[h]
\centering
\includegraphics[height=0.8\textheight]{LexisLifetableLine.pdf}
\end{figure}

# The Life Table: Age-Period-Cohort
\begin{figure}[h]
\centering
\includegraphics[height=0.8\textheight]{Lifetable}
\end{figure}

